---
title: "Impacte del COVID-19 per estat de vacunació"
subtitle: "Linear and Generalized Linear Models"
author: "Ulrik Danielsen, Alexander J Ohrt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  # html_document:
  #   code_folding: hide
  #   toc: true
  #   toc_depth: 3
  #   theme: readable
  #   highlight: textmate
  #   number_sections: true
  pdf_document:
    fig_caption: true
    number_sections: true
editor_options: 
  chunk_output_type: console
geometry:
  margin = 2cm
---

\tableofcontents
\newpage

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, warning = F, fig.width = 10)
# setwd("/home/ajo/gitRepos/glm/Deliverable")
library(dplyr)
library(lubridate)
library(ggplot2)
library(car)
```

# Introduction 

We are analyzing a data set that contains information about the impact of COVID-19 by vaccination status in 2021 in Catalunya. The data comes from various information systems of the Department of Health and the Catalan Health Service, as detailed [here](https://dadescovid.cat/documentacio). More specifically, the data shows the impact of COVID-19 (new cases, hospitalizations, critical hospitalizations and deaths) in the entirety of Catalunya, from January 1st 2021 until now, for the date of reference of each record, by gender, by age group, according to the vaccination status of the people. Keep in mind that a person is counted as many times as the number of events that has happened to the individual in question. This means that each person may be counted up to four times in the data set as a whole. More information about the data set can be found [here](https://analisi.transparenciacatalunya.cat/en/Salut/Impacte-del-COVID-19-per-estat-de-vacunaci-/6izj-g3sb). Note that the data we are using was downloaded December 17th at 12:00. 

The data set contains 6 columns. They are explained shortly below. 

| Variable | SEXE | EDAT | DATA | ESDEVENIMENT | PAUTA | RECOMPTE |
|---|---|---|---|:---|---|---|
| Explanation | Sex | Age Group | Date | Event (Severity of illness) | Vaccination status | Count |
| Factor Levels | Homme, Dona | 0-9, 10-19, 20-29, <br>30-39, 40-49, 50-59, <br>60-69, 70-79, 80-,<br>Not classified |  | Cas, Hospitalització, Critics, Defunció  | No iniciada, Parcial, Completa |  |

# Objective

The objective of the analysis is to answer the following question: 
*Do differences in the severity of COVID-19 infection exist between people that are unvaccinated, people that only have received one dose and people that are fully vaccinated?*
More precisely, does the amount of people with a *critical* infection depend on their respective vaccination statuses? 

The analysis will be performed separately for men and women. Moreover, we will only consider people that are older than 30 and only data from May 1st to December 12th. Additionally, we will only analyze differences in people that are reported with a critical event, disregarding people with new cases, people that are hospitalized (not critical) and people that have died. 

# Data Preparation

First we have to remove all people younger than 30 years old and cases before May 1st. The data is prepared according to the description above. 

```{r}
data <- read.csv("Impacte_del_COVID-19.csv") # Data updated 17.12.21.

clean <- function(data){
  data$DATA <- as.Date(data$DATA, format = "%d/%m/%Y")
  data$EDAT <- as.factor(data$EDAT)
  covid.data <- data %>% filter(DATA >= "2021-05-01" & DATA < "2021-12-13") %>% 
        filter(EDAT %in% c("30 a 39", "40 a 49", "50 a 59", "60 a 69", "70 a 79", "80 o més"))
        # People with EDAT "No classificat" are not added 
        # (these are the same people as "No classificat" in SEXE).
  
  covid.data$PAUTA <- factor(covid.data$PAUTA)
  covid.data$SEXE <- factor(covid.data$SEXE)
  
  # Create week and month variables.
  covid.data$Week <- as.factor(strftime(covid.data$DATA, format = "%W"))
  covid.data$Month <- as.factor(strftime(covid.data$DATA, format = "%m"))
  
  return(covid.data)
}
covid.data <- clean(data) %>% filter(ESDEVENIMENT == "Crítics")
```

Note also that there are `r dim(data %>% filter(SEXE == "No classificat" & EDAT == "No classificat"))[[1]]` registries in the data set that are not classified into groups in `SEXE` and `EDAT`. These rows are disregarded in the analysis. 

# Exploratory Data Analysis 

```{r}
covid.data.women <- covid.data %>% filter(SEXE == "Dona")
covid.data.men <- covid.data %>% filter(SEXE == "Home")
```

The data set is split into men and women. There are `r sum(covid.data.men$RECOMPTE)` critical cases for men and `r sum(covid.data.women$RECOMPTE)` critical cases for women in the data. 

## Critical Infections by Age

```{r}
count.in.age.group <- function(data){
  data %>% group_by(SEXE, EDAT) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}

all <- count.in.age.group(covid.data)
ggplot(all, aes(x = EDAT, y = Count)) + 
  labs(x = "Age Group", y = "Count") +
  geom_col(aes(fill = SEXE), position = "dodge") +
  geom_text(aes(label=Count, group= SEXE), vjust = 1.6, color="black", 
            size=3.5, position = position_dodge(width = 1)) +
  ggtitle("Cases of Critical COVID-19 Infection Grouped by Age") +
  theme_minimal()
```


## Evolution of Critical Infections with Time 

```{r}
count.each.week <- function(data){
  data %>% group_by(SEXE, Week) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}

week.data <- count.each.week(covid.data)
ggplot(data = week.data, aes(x = as.factor(Week), y = Count)) +
  labs(x = "Week", y = "Count") +
  geom_col(aes(fill = SEXE), position = "dodge") +
  geom_text(aes(label=Count, group= SEXE), vjust = 1.6, color="black", 
            size=2, position = position_dodge(width = 1)) +
  ggtitle("Evolution of Critical Infections by Week") +
  theme_minimal()

count.each.month <- function(data){
  data %>% group_by(SEXE, Month) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}

month.data <- count.each.month(covid.data)
ggplot(data = month.data, aes(x = as.factor(Month), y = Count)) +
  labs(x = "Month", y = "Count") +
  geom_col(aes(fill = SEXE), position = "dodge") +
  geom_text(aes(label=Count, group= SEXE), vjust = 1.6, color="black", 
            size=3, position = position_dodge(width = 0.9)) +
  ggtitle("Evolution of Critical Infections by Month") +
  theme_minimal()
```

## Critical Infections by Vaccination Status and Age 

```{r, warning=FALSE}
group.by.vacc.age.gender <- function(data){
  data %>% group_by(SEXE, EDAT, PAUTA) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}
d5 <- group.by.vacc.age.gender(covid.data)

ggplot(d5, aes(x = EDAT, y = Count)) + 
  labs(x = "Age Group", y = "Count") +
  geom_col(aes(fill = PAUTA), position = "dodge") +
  geom_text(aes(label=Count, group= PAUTA), vjust = 1.6, color="black", 
            size=2.5, position = position_dodge(width = 0.9)) +
  ggtitle("Critical Infections per Age Group based on Vaccination Status") +
  facet_wrap(~ SEXE) + 
  theme_minimal()

group.by.vacc.age <- function(data){
  data %>% group_by(EDAT, PAUTA) %>% summarise(Count = sum(RECOMPTE), .groups = "keep") 
}
d4 <- group.by.vacc.age(covid.data.women)
knitr::kable(d4, caption = "Women with Critical Infection as 
             Function of Vaccination Status and Age")
d3 <- group.by.vacc.age(covid.data.men)
knitr::kable(d3, caption = "Men with Critical Infection as 
             Function of Vaccination Status and Age")
```

## Cases by Vaccination Status per Age Group

When considering the critical cases by vaccination status and age for each gender, it is good to keep all cases per vaccination status per age group in the entire data set in mind. This data is shown below, where all four different classifications of severity of infection are counted in the data set. 

```{r}
vacc.status.per.age <- group.by.vacc.age.gender(clean(data))
ggplot(vacc.status.per.age, aes(x = EDAT, y = Count)) + 
  labs(x = "Age Group", y = "Count") +
  geom_col(aes(fill = PAUTA), position = "dodge") +
  geom_text(aes(label=Count, group= PAUTA), vjust = 1.6, color="black", 
            size=2.5, position = position_dodge(width = 0.9)) +
  ggtitle("All Cases per Vaccination Status per Age Group") +
  facet_wrap(~ SEXE) + 
  theme_minimal()
```

## Critical Cases Relative to all Cases by Vaccination Status per Age Group

The next plot shows the critical cases of COVID-19 relative to all cases by vaccination status in the entire data set per age group. Essentially, this shows the proportion of the counts of the two earlier plots. Thus, for example, even though plot 4.3 shows that there are more critical cases for completely vaccinated individuals in the ages greater than 70 (except for women aged 70 to 79), plot 4.5 shows that the proportion of cases is still greatest for the non-vaccinated (except for women aged 80 or more). This is because most of the population in these ages are in fact completely vaccinated. 

```{r}
d6 <- d5
d6$Count <- d5$Count/vacc.status.per.age$Count
ggplot(d6, aes(x = EDAT, y = Count)) + 
  labs(x = "Age Group", y = "Proportion") +
  geom_col(aes(fill = PAUTA), position = "dodge") +
  geom_text(aes(label=round(Count,4), group= PAUTA), vjust = 1.6, color="black", 
            size=2, position = position_dodge(width = 0.9)) +
  ggtitle("Proportion of Critical Cases to all Cases by Vaccination Status per Age Group") +
  facet_wrap(~ SEXE) + 
  theme_minimal()
```

## Cases by Vaccination Status with Time (Months)

All cases per vaccination status in the entire data set is shown by month in the plot below. It is apparent that the later in the year, the more cases are registered for the fully vaccinated. This is most likely a result of the fact that more and more people are getting vaccinated. i.e. the proportion of vaccinated people in the population is growing throughout the year. 

```{r}
vacc.status.with.time <- function(data){
  data %>% group_by(SEXE, PAUTA, Month) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}
vacc.with.time <- vacc.status.with.time(clean(data))
ggplot(data = vacc.with.time, aes(x = as.factor(Month), y = Count)) +
  labs(x = "Month", y = "Count") +
  geom_col(aes(fill = PAUTA), position = "dodge") +
  geom_text(aes(label=Count, group= PAUTA), vjust = 1.6, color="black",
             size=2, position = position_dodge(width = 0.9)) +
  ggtitle("Evolution of all Cases per Vaccine Status by Month") +
  facet_wrap(~ SEXE) + 
  theme_minimal()
```

## Critical Cases by Vaccination Status with Time (Months)

Only the critical cases per vaccination status is shown by month below. 

```{r}
vacc.status.with.time <- function(data){
  data %>% group_by(SEXE, PAUTA, Month) %>% summarise(Count = sum(RECOMPTE), .groups = "keep")
}
vacc.with.time.critical <- vacc.status.with.time(covid.data)
ggplot(data = vacc.with.time.critical, aes(x = as.factor(Month), y = Count)) +
  labs(x = "Month", y = "Count") +
  geom_col(aes(fill = PAUTA), position = "dodge") +
  geom_text(aes(label=Count, group= PAUTA), vjust = 1.6, color="black",
             size=2, position = position_dodge(width = 0.9)) +
  ggtitle("Evolution of Critical Cases per Vaccine Status by Month") +
  facet_wrap(~ SEXE) + 
  theme_minimal()
```

Already, the exploratory data analysis hints at the fact that the amount of critical infections does depend on the vaccination status. This should be proven statistically however, an issue that is discussed in the following. 

# Model Fit  

We wish to examine how the vaccination status affects the number of critical patients
in this period. Since we observe that the number of critical patients vary 
very much from month to month, we choose to aggregate the number of 
patients each month, and include it as a categorical covariate in our model.

```{r}
covid.data.aggr <- aggregate(RECOMPTE ~ EDAT + SEXE + PAUTA + Month,
                             covid.data, FUN = sum)
head(covid.data.aggr)
dim(covid.data.aggr)
```

## Linear Model

First we try a simple linear model with all covariates included.
As the adjusted R-squared is low, and the residuals do not fit the normal 
assumption, we reject this model.

```{r}
model1 <- lm(RECOMPTE ~ EDAT + SEXE + PAUTA + Month, data = covid.data.aggr)
summary(model1)
par(mfrow = c(2,2))
plot(model1, ask = F)
```

Since we have aggregated the number of critical patients by month, we can interpret 
this data collection as a counting process which starts each month.

## Generalized Linear Model

The described counting process is a poisson process, which we can fit as a generalized linear model
from the poisson family.
The model is fitted without interaction, and includes all covariates.
The residuals still look bad, and the model has a dispersion parameter above 
seven. 

```{r}
model2 <- glm(RECOMPTE ~ EDAT + SEXE + PAUTA + Month, data = covid.data.aggr, 
              family = poisson(link = "log"))
summary(model2)
par(mfrow = c(2,2))
plot(model2, ask = F)
# Checking for overdispersion.
PS <- sum(residuals(model2, type="pearson")^2)
PS
phi <- PS / model2$df.res
phi
```

---

Next we include an interaction between the vaccination status and month.
This makes sense because the vaccination status should become more skewed towards complete as time goes
and more people are vaccinated.

```{r}
model3 <- glm(RECOMPTE ~ EDAT + SEXE + PAUTA + Month + PAUTA:Month, data = covid.data.aggr, 
              family = poisson(link = "log"))
summary(model3)
par(mfrow = c(2,2))
plot(model3, ask = F)
# Checking for overdispersion.
PS <- sum(residuals(model3, type="pearson")^2)
PS
phi <- PS / model3$df.res
phi
```

The model is improved, but the residuals still show clear patterns, and the model
is overdispersed.

---

Now we include an interaction between vaccination status and age group.
This makes sense since the older age group likely was vaccinated earlier than
the younger age group. 

```{r}
model4 <- glm(RECOMPTE ~ EDAT + SEXE + PAUTA + Month + PAUTA:Month + EDAT:PAUTA, data = covid.data.aggr, 
              family = poisson(link = "log"))
summary(model4)
par(mfrow = c(2,2))
plot(model4, ask = F)
# Checking for overdispersion
PS <- sum(residuals(model4, type="pearson")^2)
PS
phi <- PS / model4$df.res
phi
Anova(model4)
```

Now the model shows clear improvement, and the dispersion parameter is reduced
to right above two.

---

Including an interaction between all three covariates provides us with an 
even better model, as shown below.
The graphical plots are much better, and the dispersion parameter 1.36, which 
is still slightly overdispersed, but better.
The Pearson residual for the model is 130.4, which is slightly below 
the critical value for the $\chi_{n - p}^2$ distribution. 
Thus, we can conclude that this model is a good fit to the data.


```{r}
model5 <- glm(RECOMPTE ~ SEXE + EDAT*PAUTA*Month,
              data = covid.data.aggr, 
              family = poisson(link = "log"))
summary(model5)
par(mfrow = c(2,2))
plot(model5, ask = F)
# Checking for overdispersion
PS <- sum(residuals(model5, type="pearson")^2)
PS
phi <- PS / model5$df.res
phi
Anova(model5)
qchisq(0.99, model5$df.res, lower.tail=TRUE)
```

Also, note that the AIC decreases for each of the increasingly complicated generalized linear models that are fitted to the data. This shows that the increasingly more complicated models perform better compared to the earlier models. Moreover, the Pearson residuals decrease as well, which points to the same conclusion. 

# Conclusions

The data analysed in this report contains covid status of patients in 
Catalunya between 01.05.2021 and 12.12.2021. All critical patients 
are counted each month, and are modelled as a general linear model with 
Poisson response. The covariates gender, vaccination status, age group and month
are included with interactions, and the model proved to fit the data well.

Returning to the main question: does the amount of people with a *critical* infection depend on their respective vaccination statuses? Based on the final, very complicated model, the answer is yes. The Anova table shows that all covariates containing `PAUTA` are significant to any reasonable significance level. Moreover, looking at the $p$-values from the summary of the final model, we can conclude the following: To a significance level of $\alpha = 0.05$


* `PAUTA: No iniciada` is significantly different from the baseline (`PAUTA: Completa`) ($p = 8.26\cdot10^{-5}$). The estimated parameter value is $\approx 2.82$, which means that the amount of critical infections per month is increased when individuals have not received any vaccines, compared to when individuals are completely vaccinated, when the rest of the individual's profile is identical (the other covariates are kept fixed).
* `Pauta: Parcial` is significantly different from the baseline (`PAUTA: Completa`) ($p \approx 0.27$). The estimated parameter value is $\approx 3.93$, which means that the amount of critical infections per month is increased when individuals are partially vaccinated, compared to when individuals are completely vaccinated, when the rest of the individual's profile is identical. 
* The interactions between `EDAT: 80 o més` and `PAUTA: Parcial`, and `EDAT: 80 o més` and `PAUTA: No iniciada`, are statistically significant to $\alpha$. Both parameter estimates are negative, which means that the model predicts the amount of critical infections in this age group to be larger when the individuals are fully vaccinated. This is a strange result, which goes against what one might think. When reflecting on this result, keep in mind the fact that most of the people in this age group are fully vaccinated, as shown in the exploratory data analysis. This might give rise to this slightly strange result. 
* The interaction effect between `PAUTA: Parcial` and `Month` is statistically significant from month 8 onwards. All the parameter estimates in question are negative, which means that the model predicts a smaller amount of critical infections in the later months when the individuals are partially vaccinated, compared to when the individuals are completely vaccinated. This is the case when the rest of the patient profile is kept fixed. This is also a strange result, which might depend on different relationships in the data. Plot 4.6 might help to explain this, since a larger and larger proportion of the population are vaccinated towards the end of the year. Interestingly, month 8 is the first month where the amount of cases (taking all cases into account), for both sexes, is larger for the fully vaccinated compared to the two other vaccination statuses. 

Without being sure about the reasons for all the predictions listed above, noting that some of them are strange, we can conclude that the amount of people with a critical infection depends on the vaccination status. 
