---
title: "Linear Models, Problem 4"
author: "alexaoh"
date: '`r format(Sys.Date(),"%e %B, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", eval = TRUE)

library(lattice)
library(car)
library(gridExtra)
library(grid)
library(MASS)
library(survival)
library(multcomp)
library(mvtnorm)
library(tables)
library(HH)
```

```{r}
data <- read.csv2("dcrown.csv")
data$RP <- data$PB/data$PT # Add the given ratio to the data. 
head(data)
dim(data)
summary(data)
```

### Consider Two Different Linear Models

```{r}
modA <- lm(DCrown~PT+RP+HT+A, data = data)
summary(modA)
modB <- lm(log(DCrown)~log(PT)+log(RP)+log(HT)+log(A), data = data)
summary(modB)
```
### a) Questions about `modA`

(1) All the coefficient estimates are not significantly different from zero, with a significance level of 5\%. Only `PT` and `A` (plus the intercept) are significant to this level. 

(2) The estimation of the residual variance is $(\text{Residual standard error})^2$ = `r summary(modA)$sigma^2`. 

(3) The studentized residuals are plotted below.

```{r}
plot(rstudent(modA))
abline(h = c(-2, 2), lty = 2)
```

As is apparent, 15 points are outside the interval (-2,2). This represents a percentage of approximately `r round(15*100/dim(data)[1], 2)` \% of the points, which is reasonable, considering $1.96 \approx 2$, where 1.96 is the 0.975 quantile of the standard normal distribution. Hence, (-2,2) is an approximation of a 5\% confidence interval. 


### b) Questions about `modB`

(1) In the second model all the coefficient estimates are significantly different from zero, with a significance level of 5\%. 

(2) The estimation of the residual variance is $(\text{Residual standard error})^2$ = `r summary(modB)$sigma^2`. 

(3) The studentized residuals are plotted below.

```{r}
plot(rstudent(modB))
abline(h = c(-2, 2), lty = 2)
```

As is apparent, 11 points are outside the interval (-2,2). This represents a percentage of approximately `r round(11*100/dim(data)[1], 2)`\% of the points, which is reasonable, considering $1.96 \approx 2$, where 1.96 is the 0.975 quantile of the standard normal distribution. Hence, (-2,2) is an approximation of a 5\% confidence interval. 

### c)

I would choose `modB` since, compared to `modB`, $R^2$ is larger, the F-statistic is larger (even though both have small $p$-values for the F-test) and all the coefficients are significant (in contrast to the first model, which only has a few estimates that are significantly non-zero, to a 5\% level). Note that I have assumed that the assumptions of the linear model are verified for both models (as it says in the problem), which is why these assumptions are not discussed.  

### d)

```{r}
new.data <- data.frame(PT = 0.4, PB = 0.6, HT = 2.3, A = 10, RP = 0.6/0.4)
predict(modB, new.data, interval = "confidence", level = 0.95)
```

