---
title: "Linear Models, Problem 4"
author: "alexaoh"
date: '`r format(Sys.Date(),"%e %B, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", eval = TRUE)

library(lattice)
library(car)
library(gridExtra)
library(grid)
library(MASS)
library(survival)
library(multcomp)
library(mvtnorm)
library(tables)
library(HH)
```

```{r}
data <- read.csv2("dcrown.csv")
data$RP <- data$PB/data$PT # Add the given ratio to the data. 
head(data)
dim(data)
summary(data)
```

### Consider Two Different Linear Models

```{r}
modA <- lm(DCrown~PT+RP+HT+A, data = data)
summary(modA)
modB <- lm(log(DCrown)~log(PT)+log(RP)+log(HT)+log(A), data = data)
summary(modB)
```
### a) Questions about `modA`

(1) All the coefficient estimates are not significantly different from zero, with a significance level of 5\%. Only `PT` and `A` (plus the intercept) are significant to this level. 

(2) The estimation of the residual variance is $(\text{Residual standard error})^2$ = `r summary(modA)$sigma^2`. 

(3) The studentized residuals are plotted below.

```{r}
# Plot studentized residuals. 
```


### b) Questions about `modB`

